Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-09-01 interimreport2.Rnw
Figures and cost benefit changes
Large changes to sweave code, especially relating to figures, and
altered c/b ratio analyses after going through with Tim

\title{Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\maketitle

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations. We provide preliminary cost--benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:
\begin{itemize}
\item text.
\item text.
\item text.
\end{itemize}

@@

As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission. A brief review of the history of measles in New Zealand was provided in our previous interim report. In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the cost--benefit analyses for vaccination dependent on differing scenarios.

@@

\subsection{Risk analyses methods}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist.

The University of Otago denominator data provided were not to the same detail as the ESR case data. Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories. The ethnicity denominator data was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes. After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mfrow=c(1,1))
hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F)
hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n")
@
\end{center}
\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

@@

This large number of categories, some with population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:cases}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:cases}.

\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(ttyr, file="", table.env=FALSE,rowname=NULL)
@
\label{table:cases}
\end{table}

@@

Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over- or under-represented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{fig:}), leading to very large standard error in regression analyses.

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="Category")
legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
       col=c(1:4,6),pch=16,bty="n")
@
\end{center}
\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

@@

However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes, and there several issues with estimating the denominator data for these group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007.

For all our statistical analyses (including those above not shown) we used a Poisson error strutuce, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significat interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}
Where $y$ = cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.

@@

The distribution of the cases per category used are in Figure~\fig{histcasecat}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20)
@
\end{center}
\caption{Histogram of cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

%\begin{figure}[h!]
%\begin{center}
%<<fig=TRUE,echo=FALSE,width = 11>>=
%plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
%@
%\end{center}
%\caption{test}
%\label{fig:test}
%\end{figure}
The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width =11, height=11>>=
res<-predict(model3)
plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey")
@
\end{center}
\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
\label{fig:predict}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
hist(model3$residuals,main="",xlab="residuals",col="grey")
@
\end{center}
\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
\label{fig:resid}
\end{figure}
The significance of the different variables can be seen here, in the ANOVA results:

@@

The results of the model suggest that, apart from over representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age. This can be seen in Figure~\ref{fig:ageinyears}.

People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.

In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected. The very young are still most likely to be infections, but older teenagers are more likely to be represented (Figure~\ref{fig:ageinyears}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly these results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.

@@

The analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at risk. In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5, and European 6--17 year old children. The results of these analyses are discussed in the cost--benefit section below (Section\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR). A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR. This work is ongoing.

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age
\item Pacific people are more at risk on a per capita basis
\item There is some support for those living in better socio-economic situations being at greater risk of measles
\item There is support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
\end{itemize}

@@

In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the cost--benefit analysis (Section~\ref{sub:cost_benefit}).

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
plot(pop,xlab="Age",ylab="Population")
points(naive,pch=16)
legend("topright",c("Population","Naive"),pch=c(1,16),bty="n")
@
\end{center}
\caption{New Zealand population by age and estimates numbers of naive people in each age class}
\label{fig:naive}
\end{figure}

@@

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
     \label{fig:r0}
\end{figure}

@@

We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals. These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the cost--benefit analyses section below (Section~\ref{sub:cost_benefit}).

@@

The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report. The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one. As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious. However, the results of these differences are presented in the cost--benefit analysis section (Section~\ref{sub:cost_benefit}).

These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic and we explore the cost--benefit effects of this in Section~\ref{sub:cost_benefit}.

@@

To use the results from our modeling exercise to help inform the appropriate measles vaccination coverage. Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}). These figures can be reached in a number of different ways, and these are discussed in the cost--benefit section (Section~\ref{sub:cost_benefit}).

\subsection{Summary of modelling}
These results suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}). However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary. In Ireasl researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached. Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate. 
In summary:
\begin{itemize}
\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
 \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
 \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0 to 53\%, depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
\end{itemize}

@@

\subsection{Cost analyses results}
\label{sub:cost_results}
Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}. The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.

@@

<<echo=FALSE>>=
hosp<-read.csv("hospital.csv",header=T)
library(ggplot2)
p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")
@

@@

The results in the two tables show two alternative trends. In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}. However, if $R_v$ is less than one and the previous recent history of cases since 1997 is the 'status quo' then the additional effort to vaccinate the naive population based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r_0} is not a cost effective exercise (Table~\ref{tab:cost_mean}.

>>>>

\title{Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\maketitle

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations. We provide preliminary benefit--cost analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:
\begin{itemize}
\item Age is the best predictor of risk of measles infection in multi-variate regression analyses, though some groups such as Pacific people, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is very beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
\end{itemize}

@@

As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission. A brief review of the history of measles in New Zealand was provided in our previous interim report. In this report were present regression analyses to determine which populations are most at risk, the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the benefit--cost analyses for vaccination dependent on differing scenarios.

@@

\subsection{Risk analyses methods}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist among cases allowing better understanding of risk of measles infection.

The University of Otago denominator data provided were not to the same detail as the ESR case data. Notably, the denominator age data were categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories. The ethnicity denominator data were not Prioritized Ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes. After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mfrow=c(1,1))
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
@
\end{center}
\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

@@

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: 1--5 and 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:case}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(ttyr, file="", table.env=FALSE,rowname=NULL)
@
\label{table:case}
\end{table}

@@

The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="Category",cex.lab=2)
legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
@
\end{center}
\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

@@

However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.

@@

The distribution of the cases per category used are in Figure~\fig{fig:histcasecat}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
@
\end{center}
\caption{Histogram of cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width =11, height=11>>=
res<-predict(model3)
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
@
\end{center}
\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
\label{fig:predict}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
@
\end{center}
\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
\label{fig:resid}
\end{figure}
The significance of the different predictor variables can be seen here, in the ANOVA results:

@@

The results of the model suggest that, apart from over-representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).

People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.

In later outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young are still most likely to be infections, but older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.

@@

The analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at risk. In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European 6--17 year old children. The results of these analyses are discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR). A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age
\item Pacific people are statistically more at risk on a per capita basis
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
\item There is some statistical support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
\end{itemize}

@@

In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the benefit--cost analysis (Section~\ref{sub:cost_benefit}).

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
pop<-read.csv("popnsize.csv",header=T)
colnames(pop)<-0:100
pop<-t(pop)
impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
#length(pop)
naive<-round(pop-(pop*impop))
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
plot(pop,xlab="Age",ylab="Population",cex.lab=2)
points(naive,pch=16)
legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
@
\end{center}
\caption{New Zealand population by age and estimates numbers of naive people in each age class}
\label{fig:naive}
\end{figure}

@@

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks}}
     \label{fig:r0}
\end{figure}

@@

We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals. These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).

@@

The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report. The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one. As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious. However, the results of these differences are presented in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).

These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, and we explore the benefit--cost effects of this in Section~\ref{sub:cost_benefit}.

@@

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage. Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}). These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).

\subsection{Summary of modelling}
These results suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}). However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary. In Isreal researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached. Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate. 
In summary:
\begin{itemize}
\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
 \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
 \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
\end{itemize}

@@

\subsection{Cost analyses results}
\label{sub:cost_benefit}
Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}. The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.

@@

<<echo=FALSE>>=
hosp<-read.csv("hospital.csv",header=T)
library(ggplot2)
p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
@

@@

The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}. The results in the two tables show two alternative trends. In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}). However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
